Categories of Optics

### 2018-06-12 paper.tex
Fill in some little bits

@@

\newcommand{\rep}[2]{{\ensuremath \langle #1 \mid #2 \rangle}}
\newcommand{\repthree}[3]{{\ensuremath \langle #1 \mid #2 \mid #3 \rangle}}
\newcommand{\repfour}[4]{{\ensuremath \langle #1 \mid #2 \mid #3 \mid #4 \rangle}}

@@

Graphically, this is placing the first optic into the hole of the second:
\begin{center}
  \input{diagrams/generic-optic-noline.tikz}
  \qquad\raisebox{1.5cm}{$\circ$}\qquad
  \input{diagrams/generic-optic2-noline.tikz} \\
  \raisebox{1.5cm}{$:=$}\qquad
  \input{diagrams/optic-composition.tikz}
\end{center}

The identity optic $(S, S') \hto (S, S')$ is given by $\rep{\lambda^{-1}_S}{\lambda_{S'}}$, where $\lambda_S : I \otimes S \to S$ is the left unitor for $S$:
\begin{center}
  \input{diagrams/identity-optic-full.tikz}
\end{center}
This dashed line above the diagram represents the unit object. As is standard, we will elide the unit object and unitors in future diagrams unless they are necessary. The diagram for the identity becomes:
\begin{center}
  \input{diagrams/identity-optic.tikz}
\end{center}

\begin{proposition}
  The above data form a category $\Optic_\C$.
\end{proposition}
\begin{proof}
  In~\cite[Section 6]{Doubles} this is proven abstractly, by exhibiting this category as the Kleisli category for a monad in the bicategory $\Prof$. Here we prove the properties directly.

@@

\begin{proposition}
  There is a functor $\iota : \C \times \C^\op \to \Optic_\C$, which is given on objects by $\iota(S, S') = (S, S')$ and for a morphism $(f, g) : (S, S') \to (A, A')$ by $\iota(f, g) = \rep{ \lambda_A^{-1} f}{ g \lambda_{A'}}$.
\end{proposition}
\begin{proof}
  Graphically, this is:
  \begin{center}
    \input{diagrams/iota.tikz}
  \end{center}

  This preserves identities, as the identity on an object $(S, S')$ in $\Optic_\M$ is defined to be exactly $\rep{ \lambda^{-1}_S}{ \lambda_{S'}}$. To check functoriality, suppose we have $(f, g) : (S, S') \to (A, A')$ and $(f', g') : (A, A') \to (B, B')$ in $\C \times \C^\op$. Then:
  \begin{align*}
    \iota(f', g') \circ \iota(f, g)
    &= \rep{ \lambda^{-1}_B f'}{ g' \lambda_{B'}} \circ \rep{ \lambda^{-1}_A f}{ g \lambda_{A'}} \\
    &= \rep{ (I\otimes (\lambda^{-1}_B f'))\lambda^{-1}_A f}{ g \lambda_{A'} (I\otimes (g' \lambda_{B'})} && \text{(By definition of $\circ$)}\\
    &= \rep{ (I \otimes \lambda^{-1}_B) (I \otimes f')\lambda^{-1}_A f}{ g \lambda_{A'} (I \otimes g')(I\otimes \lambda_{B'}} && \text{(Functoriality of $I \otimes -$)}\\
    &= \rep{ (I\otimes \lambda^{-1}_B) \lambda^{-1}_B f' f}{ g g' \lambda_{B'} (I\otimes \lambda_{B'}} && \text{(Naturality of $\lambda$)}\\
    &= \rep{ (\lambda^{-1}_I \otimes B) \lambda^{-1}_B f' f}{ g g' \lambda_{B'} (\lambda_I \otimes B'} && \text{(Unitality of action)} \\
    &= \rep{ \lambda^{-1}_B f' f}{ g g' \lambda_{B'} (\lambda_I \otimes B') (\lambda^{-1}_I \otimes B')} && \text{(Coend relation)}  \\
    &= \rep{ \lambda^{-1}_B f'f}{ g g' \lambda_{B'}} \\
    &= \iota(f'f, gg')
  \end{align*}
  Graphically, there is not much to do:
  \begin{center}
    \input{diagrams/iota-functorial-left.tikz}
    \qquad \raisebox{0.3cm}{$=$} \qquad
    \input{diagrams/iota-functorial-right.tikz}
  \end{center}

@@

\begin{theorem}
  $\Optic_\C$ is symmetric monoidal, where $(S, S') \otimes (T, T') = (S \otimes T, S' \otimes T')$, the unit object is $(I, I)$, and the action on a pair of morphisms $\rep{ l}{ r} : S \hto A$ and $\rep{ l'}{ r'} : T \hto B$ is given by:
  \begin{center}
    \input{diagrams/tensor-on-morphisms.tikz}
  \end{center}
\end{theorem}
\begin{proof}
  Equationally, suppose the two optics have complements $M$ and $N$ respectively. Then:
  \begin{align*}
    \rep{ l}{ r} \otimes \rep{ l'}{ r'} &:= \rep{(M \otimes s_{A,N} \otimes B)(l \otimes l')}{(r \otimes r')(M \otimes s_{A',N} \otimes B')}
  \end{align*}
  This is well-defined, as the calculus of string diagrams allows us to combine the coend relation on each optic to establish the coend relation for the compound diagram. \todo{draw this?}

  Let us check functoriality of $\otimes$. Suppose we have optics:
  \begin{align*}
    \rep{ l_1}{ r_1} : (S_1, S_1') &\hto (S_2, S_2') \\
    \rep{ l_2}{ r_2} : (S_2, S_2') &\hto (S_3, S_3') \\
    \rep{ p_1}{ q_1} : (T_1, T_1') &\hto (T_2, T_2') \\
    \rep{ p_2}{ q_2} : (T_2, T_2') &\hto (T_3, T_3')
  \end{align*}
  The string diagram for $(\rep{ l_2}{ r_2} \circ \rep{ l_1}{ r_1}) \otimes (\rep{ p_2}{ q_2} \circ \rep{ p_1}{ q_1})$ is:
  \begin{center}
    \input{diagrams/tensor-functorial-left.tikz}
  \end{center}
  And for $(\rep{ l_2}{ r_2} \otimes \rep{ p_2}{ q_2}) \circ (\rep{ l_1}{ r_1} \otimes \rep{ p_1}{ q_1})$ is:
  \begin{center}
    \input{diagrams/tensor-functorial-right.tikz}
  \end{center}
  These two diagrams are equivalent. We can use the naturality of the symmetry morphism to push $l_2$ and $r_2$ through to be next to $l_1$ and $r_1$ respectively. This creates two extra twists that can be cancelled in the center of the diagram.

@@

  Note that because $\iota(S, S') = (S, S')$, required equations for $\iota$ to be a monoidal functor hold by definition (although we don't yet know that $\Optic_\otimes$ is monoidal). The pentagon and triangle axioms then hold in $\Optic_\otimes$, as they are the image of the same axioms in $\C \times \C^\op$ under $\iota$. The only remaining thing to verify is that these structure maps are natural in $\Optic_\otimes$.

  Let us check that $\alpha_{(R, R'), (S, S'), (T, T')}$ is natural in $(R, R')$. Suppose $f : (Q, Q') \hto (R, R')$ has representative $(l, r) : (Q \to M \otimes R, M \otimes R' \to Q')$, then:
  \begin{align*}
    &\alpha_{(R, R'), (S, S'), (T, T')} \circ (f \otimes (S, S')) \otimes (T, T') \\
    &= \iota(\alpha_{R,S,T}, \alpha_{R',S',T'}^{-1})  \circ (\rep{ l}{ r} \otimes (S, S')) \otimes (T, T') \\
    &= \rep{\lambda_{R \otimes (S \otimes T)}\alpha_{R,S,T}}{\alpha_{R',S',T'}^{-1} \lambda^{-1}_{(R' \otimes S') \otimes T'}}  \circ \rep{(l \otimes S) \otimes T}{ (r \otimes S') \otimes T'} \\
    &= \rep{(M \otimes \lambda_{R \otimes (S \otimes T)}\alpha_{R,S,T})((l \otimes S) \otimes T)}{((r \otimes S') \otimes T') (M \otimes \alpha_{R',S',T'}^{-1} \lambda^{-1}_{(R' \otimes S') \otimes T'} )} \\
    &=\rep{(M \otimes \alpha_{R,S,T})((l \otimes S) \otimes T)}{((r \otimes S') \otimes T') (M \otimes \alpha_{R',S',T'}^{-1} )} \\
    &= \rep{(l \otimes (S \otimes T))\alpha_{Q,S,T}}{\alpha_{Q',S',T'}^{-1}(r \otimes (S' \otimes T'))} \\
    &= \rep{(I \otimes l \otimes (S \otimes T)) \lambda_{Q \otimes (S \otimes T)}\alpha_{Q,S,T}}{(I \times r \otimes (S' \otimes T')) \alpha_{Q',S',T'}^{-1} \lambda^{-1}_{Q' \otimes (S' \otimes T')}} \\
    &= \rep{l \otimes (S \otimes T)}{r \otimes (S' \otimes T')} \circ \rep{\lambda_{Q \otimes (S \otimes T)}\alpha_{Q,S,T}}{\alpha_{Q',S',T'}^{-1} \lambda^{-1}_{Q' \otimes (S' \otimes T')} } \\
    &= \rep{ l}{ r} \otimes ((S, S') \otimes (T, T')) \circ \iota(\alpha_{Q,S,T}, \alpha_{Q',S',T'}^{-1}) \\
    &= f \otimes ((S, S') \otimes (T, T')) \circ \alpha_{(Q, Q'), (S, S'), (T, T')}
  \end{align*}
  \todo{There is some cheating going on here, composition actually adds an extra $\alpha$, but that wouldn't get in the way.}

  \todo{convert this to a string diagram}
\end{proof}

\begin{remark}
  In the Haskell \texttt{lens} library, the monoidal product on optics is denoted ``\texttt{alongside}'' for the product and ``\texttt{without}'' for the coproduct.
\end{remark}

@@

\begin{proposition}
  $\iota : \C \times \C^\op \to \Optic_\C$ is a monoidal functor.
\end{proposition}
\begin{proof}
  This is clear, as the structure morphisms in $\Optic_\C$ are exactly the structure morphisms in $\C \times \C^\op$ under $\iota$.
\end{proof}

@@

In particular, for any $S \in \C$, the image of the identity yields an optic $c_S : (S, S) \hto (I, I)$, that we will call the connector:
\begin{center}
  \input{diagrams/connector-full.tikz}
\end{center}
Or, again omitting the unitors:
\begin{center}
  \input{diagrams/connector.tikz}
\end{center}

@@

\begin{proposition}\label{prop-change-of-action-monoidal}
  A monoidal functor $F : \C \to \D$ induces a functor $\Optic(F) : \Optic_\C \to \Optic_\D$.
\end{proposition}
\begin{proof}
  On objects this is simply $\Optic(F)(S, S') = (FS, FS')$. On a morphism $\rep{ l}{ r} : (S, S') \to (A, A')$, we define
  \begin{align*}
    F\rep{ l}{ r} := \rep{\phi^{-1}_{M,A} (Fl)}{(Fr) \phi_{M,A'}}
  \end{align*}
  where $\phi_{M,A} : FM \otimes FA \to F(M \otimes A)$ is the structure map of the monoidal functor.
  \todo{functoriality}
  \todo{monoidality}
\end{proof}

\begin{proposition}\label{prop-optic-functor}
  Collectively this operation defines a 2-functor \[\Optic : \SymmMonCat \to \SymmMonCat.\] \qed
\end{proposition}
\todo{Haven't actually checked natural transformations}
\todo{Also we run into the large coend issue again, it's probably not well defined on every symm mon cat.}

@@

In this section we establish a universal property of the $\Optic$ construction. The idea is that every optic $\rep{ l}{ r} : (S, S') \hto (A, A')$ consists of a morphism $S \to M \otimes A$ and the `formal dual' of a morphism $M \otimes A' \to S'$, composed with a `formal counit' that traces out the object $M$:
\begin{center}
  \input{diagrams/generic-optic-folded.tikz}
\end{center}

We assume temporarily that $\C$ is a \emph{strict} symmetric monoidal category. It will be convenient to equip $\Optic_\C$ with a slightly different symmetric monoidal structure. The universal property for $\Optic_\C$ given in this section is an argument for this being the ``morally correct'' tensor, although it does seem a little strange.

\begin{definition}
  The \emph{switched} monoidal product on $\Optic_\C$ is given on objects by
  \begin{align*}
    (S, S') \switched (T, T') := (S \otimes T, T' \otimes S')
  \end{align*}
  And on morphisms $\rep{ l}{ r} : S \hto A$ and $\rep{ l'}{ r'} : T \hto B$ by:
  \begin{center}
    \input{diagrams/switched-tensor-on-morphisms.tikz}
  \end{center}
\end{definition}

@@

  The categories are monoidally equivalent via the identity functor $I : (\Optic_\C, \switched, (I, I)) \to (\Optic_\C, \otimes, (I, I))$, where the structure isomorphisms are given by
  \begin{align*}
    \iota(\id_{S \otimes S'}, s_{T, T'}) : I(S, S') \otimes I(T, T') \to I((S, S') \switched (T, T'))
  \end{align*}

@@

Note that because $\C_d$ is symmetric monoidal and has the same collection of objects as $\C$, the symmetric monoidal structure morphisms of $\C$ must all be in $\C_d$.

@@

\todo{\begin{example}
    The funny graph example from the coherence paper?
  \end{example}}

@@

  Note that the diagrams required to commute all terminate with the unit $I$, so in view of Proposition~\ref{prop-costates} we should not be surprised that they correspond to equality of maps in $\C$.
\end{proof}

@@

  First, $F$ preserves the dualisable morphisms. If $\iota(f, g) : (S, S') \hto (A, A')$,
  \begin{align*}
    F\iota(f, g)
    &= F(\rep{ \lambda_A^{-1} f}{ g \lambda_{A'}}) \\
    &= \rep{\phi^{-1}_{I,A} (F(\lambda_A^{-1} f))}{(F(g \lambda_{A'})) \phi_{I,A'}} \\
    &= \rep{ \phi^{-1}_{I,A} (F\lambda_A^{-1}) (Ff)}{(Fg)(F \lambda_{A'}) \phi_{I,A'}} \\
    &= \rep{ (\phi^{-1}_I \otimes FA) \phi^{-1}_{I,A} (F\lambda_A^{-1}) (Ff)}{(Fg)(F \lambda_{A'}) \phi_{I,A'} (\phi_I \otimes FA)} \\
    &= \rep{ \lambda_{FA}^{-1} (Ff)}{ (Fg)\lambda_{FA'}} \\
    &= \iota(Ff, Fg)
  \end{align*}

  And it also preserves the counits:
  \begin{align*}
    F(\varepsilon_{(S, S')})
    &= F(c_{S \otimes S'}) \\
    &= F(\rep{ \rho_S^{-1}}{ \rho_{S'}}) \\
    &= \rep{\phi^{-1}_{S,I} (F \rho_S^{-1})}{(F \rho_{S'}) \phi_{S',I}} \\
    &= \rep{ (FS \otimes \phi_I^{-1}) \phi^{-1}_{S,I} (F \rho_S^{-1})}{ (F \rho_{S'}) \phi_{S',I} (FS \otimes \phi_I)} \\
    &= \rep{ \rho_{FS}^{-1}}{ \rho_{FS'} } \\
    &= \varepsilon_{(FS, FS')}
  \end{align*}

@@

\begin{proposition}\label{prop-optic-decompose}
  Suppose $\rep{ l}{ r} : (S, S') \hto (A, A')$ has complement $M$. Then
  \begin{align*}
    \rep{ l}{ r} = ((A, I) \switched \varepsilon_{(M, I)} \switched (I, A'))(j(s_{M,A}l) \switched j(rs_{A',M})^*)
  \end{align*}
  where $j : \C \to \Optic_\C$ is the functor $j(A) := \iota(A, I)$.
  \todo{The symmetries could have been avoided in the above expression if the Optic category had been defined with the tensor in the opposite order!}
\end{proposition}
\begin{proof}
  First note that because $\C$ is strict monoidal, the counit $\varepsilon_{(M, I)} : (M \otimes I, M \otimes I) = (M, M) \hto (I, I)$ is equal to the connector $c_M : (M, M) \hto (I, I)$.

  Then, up to strictness of the monoidal unit, we are composing the two optics
  \begin{center}
    \input{diagrams/optic-decomposed-outer.tikz}
  \end{center}
  and
  \begin{center}
    \input{diagrams/optic-decomposed-inner.tikz}
  \end{center}
  so the two pairs of twists cancel, and we are left exactly with the diagram for $\rep{ l}{ r}$.
\end{proof}

\begin{theorem}
  \label{optic-is-free-teleological-cat}
  Suppose $\C$ is a strict symmetric monoidal category and $\T$ is a teleological category. For all symmetric monoidal functors $F : \C \to \T_d$, there exists a teleological functor $K : \Optic_\C \to \T$, unique up to isomorphism, such that $Kj \cong F$.
  % $\Optic$ is left adjoint to $(-)_d$, so $\Optic_\C$ is the free teleological category on a category of dualisable morphisms $\C$.
\end{theorem}
\begin{proof}
  On objects, define $K(S, S') = FS \otimes (FS')^*$. Suppose $\rep{ l}{ r} : (S, S') \hto (A, A')$ is an optic. By the previous Proposition,
  \begin{align*}
    \rep{ l}{ r} = ((A, I) \switched \varepsilon_{(M, I)} \switched (I, A'))(j(s_{M,A}l) \switched j(rs_{M,A'})^*)
  \end{align*}
  But now, in order for $K$ to be a teleological functor with $Kj \cong F$, each piece of this expression must be preserved by $K$, so we are forced to define:
  \begin{align*}
    K\rep{ l}{ r} = (FA \otimes \varepsilon_{FM} \otimes (FA')^*)(F(s_{M,A}l) \otimes (F(rs_{A', M}))^* )
  \end{align*}

>>>>

\newcommand{\lenslib}{\texttt{lens}}

@@

\newcommand{\rep}[2]{{\ensuremath \left\langle #1 \mid #2 \right\rangle}}
\newcommand{\repthree}[3]{{\ensuremath \langle #1 \mid #2 \mid #3 \rangle}}
\newcommand{\repfour}[4]{{\ensuremath \langle #1 \mid #2 \mid #3 \mid #4 \rangle}}

@@

Graphically, this is substituting the first optic for the hole of the second:
\begin{center}
  \input{diagrams/generic-optic-noline.tikz}
  \qquad\raisebox{1.5cm}{$\circ$}\qquad
  \input{diagrams/generic-optic2-noline.tikz} \\
  \raisebox{1.5cm}{$:=$}\qquad
  \input{diagrams/optic-composition.tikz}
\end{center}

The identity optic $(S, S') \hto (S, S')$ is given by $\rep{\lambda^{-1}_S}{\lambda_{S'}}$, where $\lambda_S : I \otimes S \to S$ is the left unitor for $S$:
\begin{center}
  \input{diagrams/identity-optic-full.tikz}
\end{center}
This dashed line above the diagram represents the unit object. It is common in string diagrams to omit unitors and the unit object unless they are necessary to make sense of the diagram. In this case, we would draw:
\begin{center}
  \input{diagrams/identity-optic.tikz}
\end{center}

\begin{proposition}\label{prop-optic-is-cat}
  The above data form a category $\Optic_\C$.
\end{proposition}
\begin{proof}
  In~\cite[Section 6]{Doubles} this is proven abstractly, by exhibiting this category as the Kleisli category for a monad in the bicategory $\Prof$. Here we prove the properties directly.

@@

\begin{proposition}
  There is a functor $\iota : \C \times \C^\op \to \Optic_\C$, which is given on objects by $\iota(S, S') = (S, S')$ and for a morphism $(f, g) : (S, S') \to (A, A')$ by $\iota(f, g) = \rep{\lambda_A^{-1} f}{g \lambda_{A'}}$.
\end{proposition}
\begin{proof}
  Graphically, this is:
  \begin{center}
    \input{diagrams/iota.tikz}
  \end{center}

  This preserves identities, as the identity on an object $(S, S')$ in $\Optic_\M$ is defined to be exactly $\rep{\lambda^{-1}_S}{\lambda_{S'}}$. To check functoriality, suppose we have $(f, g) : (S, S') \to (A, A')$ and $(f', g') : (A, A') \to (B, B')$ in $\C \times \C^\op$. Then:
  \begin{align*}
    \iota(f', g') \circ \iota(f, g)
    &= \rep{\lambda^{-1}_B f'}{g' \lambda_{B'}} \circ \rep{\lambda^{-1}_A f}{g \lambda_{A'}} \\
    &= \rep{(I\otimes (\lambda^{-1}_B f'))\lambda^{-1}_A f}{g \lambda_{A'} (I\otimes (g' \lambda_{B'})} && \text{(By definition of $\circ$)}\\
    &= \rep{(I \otimes \lambda^{-1}_B) (I \otimes f')\lambda^{-1}_A f}{g \lambda_{A'} (I \otimes g')(I\otimes \lambda_{B'}} && \text{(Functoriality of $I \otimes -$)}\\
    &= \rep{(I\otimes \lambda^{-1}_B) \lambda^{-1}_B f' f}{g g' \lambda_{B'} (I\otimes \lambda_{B'}} && \text{(Naturality of $\lambda$)}\\
    &= \rep{(\lambda^{-1}_I \otimes B) \lambda^{-1}_B f' f}{g g' \lambda_{B'} (\lambda_I \otimes B'} && \text{(Unitality of action)} \\
    &= \rep{\lambda^{-1}_B f' f}{g g' \lambda_{B'} (\lambda_I \otimes B') (\lambda^{-1}_I \otimes B')} && \text{(Coend relation)}  \\
    &= \rep{\lambda^{-1}_B f'f}{g g' \lambda_{B'}} \\
    &= \iota(f'f, gg')
  \end{align*}
  Graphically, there is not much to do:
  \begin{center}
    \input{diagrams/iota-functorial-left.tikz}
    \qquad \raisebox{0.3cm}{$=$} \qquad
    \input{diagrams/iota-functorial-right.tikz}
  \end{center}

@@

\begin{theorem}
  $\Optic_\C$ is symmetric monoidal, where $(S, S') \otimes (T, T') = (S \otimes T, S' \otimes T')$, the unit object is $(I, I)$, and the action on a pair of morphisms $\rep{l}{r} : S \hto A$ and $\rep{l'}{r'} : T \hto B$ is given by:
  \begin{center}
    \input{diagrams/tensor-on-morphisms.tikz}
  \end{center}
\end{theorem}
\begin{proof}
  Suppose the two optics have complements $M$ and $N$ respectively. Written equationally, their tensor is:
  \begin{align*}
    \rep{l}{r} \otimes \rep{l'}{r'} &:= \rep{(M \otimes s_{A,N} \otimes B)(l \otimes l')}{(r \otimes r')(M \otimes s_{A',N} \otimes B')}
  \end{align*}
  This is well-defined, as the calculus of string diagrams allows us to combine the coend relation on each optic to establish the coend relation for the compound diagram. \todo{draw this?}

  Let us check functoriality of $\otimes$. Suppose we have optics:
  \begin{align*}
    \rep{l_1}{r_1} : (S_1, S_1') &\hto (S_2, S_2') \\
    \rep{l_2}{r_2} : (S_2, S_2') &\hto (S_3, S_3') \\
    \rep{p_1}{q_1} : (T_1, T_1') &\hto (T_2, T_2') \\
    \rep{p_2}{q_2} : (T_2, T_2') &\hto (T_3, T_3')
  \end{align*}
  The string diagram for $(\rep{l_2}{r_2} \circ \rep{l_1}{r_1}) \otimes (\rep{p_2}{q_2} \circ \rep{p_1}{q_1})$ is:
  \begin{center}
    \input{diagrams/tensor-functorial-left.tikz}
  \end{center}
  And for $(\rep{l_2}{r_2} \otimes \rep{p_2}{q_2}) \circ (\rep{l_1}{r_1} \otimes \rep{p_1}{q_1})$ is:
  \begin{center}
    \input{diagrams/tensor-functorial-right.tikz}
  \end{center}
  These two diagrams are equivalen: We can use the naturality of the symmetry morphism to push $l_2$ and $r_2$ past the crossing to be next to $l_1$ and $r_1$ respectively. This creates two extra twists that can be cancelled in the center of the diagram (implicitly using the coend relation).

@@

  Note that because $\iota(S, S') = (S, S')$, required equations for $\iota$ to be a monoidal functor hold by definition (although we don't yet know that $\Optic_\otimes$ is monoidal). The pentagon and triangle equations then hold in $\Optic_\otimes$, as they are the image of the same axioms in $\C \times \C^\op$ under $\iota$. The only remaining thing to verify is that these structure maps are natural in $\Optic_\otimes$.

  Let us check that $\alpha_{(R, R'), (S, S'), (T, T')}$ is natural in $(R, R')$. Suppose $\rep{l}{r} : (Q, Q') \hto (R, R')$ is an optic with complement $M$.
  \begin{align*}
    &\alpha_{(R, R'), (S, S'), (T, T')} \circ (f \otimes (S, S')) \otimes (T, T') \\
    &= \iota(\alpha_{R,S,T}, \alpha_{R',S',T'}^{-1})  \circ (\rep{l}{r} \otimes (S, S')) \otimes (T, T') \\
    &= \rep{\lambda_{R \otimes (S \otimes T)}\alpha_{R,S,T}}{\alpha_{R',S',T'}^{-1} \lambda^{-1}_{(R' \otimes S') \otimes T'}}  \circ \rep{(l \otimes S) \otimes T}{(r \otimes S') \otimes T'} \\
    &= \rep{(M \otimes \lambda_{R \otimes (S \otimes T)}\alpha_{R,S,T})((l \otimes S) \otimes T)}{((r \otimes S') \otimes T') (M \otimes \alpha_{R',S',T'}^{-1} \lambda^{-1}_{(R' \otimes S') \otimes T'} )} \\
    &=\rep{(M \otimes \alpha_{R,S,T})((l \otimes S) \otimes T)}{((r \otimes S') \otimes T') (M \otimes \alpha_{R',S',T'}^{-1} )} \\
    &= \rep{(l \otimes (S \otimes T))\alpha_{Q,S,T}}{\alpha_{Q',S',T'}^{-1}(r \otimes (S' \otimes T'))} \\
    &= \rep{(I \otimes l \otimes (S \otimes T)) \lambda_{Q \otimes (S \otimes T)}\alpha_{Q,S,T}}{(I \times r \otimes (S' \otimes T')) \alpha_{Q',S',T'}^{-1} \lambda^{-1}_{Q' \otimes (S' \otimes T')}} \\
    &= \rep{l \otimes (S \otimes T)}{r \otimes (S' \otimes T')} \circ \rep{\lambda_{Q \otimes (S \otimes T)}\alpha_{Q,S,T}}{\alpha_{Q',S',T'}^{-1} \lambda^{-1}_{Q' \otimes (S' \otimes T')} } \\
    &= \rep{l}{r} \otimes ((S, S') \otimes (T, T')) \circ \iota(\alpha_{Q,S,T}, \alpha_{Q',S',T'}^{-1}) \\
    &= f \otimes ((S, S') \otimes (T, T')) \circ \alpha_{(Q, Q'), (S, S'), (T, T')}
  \end{align*}

Naturality in the other variables is similar, as is the naturality of the unitors.
  \todo{convert this to a string diagram?}
\end{proof}

\begin{remark}
  In the Haskell \lenslib{} library, the monoidal product on optics is denoted ``\texttt{alongside}'' for the product and ``\texttt{without}'' for the coproduct.
\end{remark}

@@

\begin{proposition}
  $\iota : \C \times \C^\op \to \Optic_\C$ is a strict monoidal functor.
\end{proposition}
\begin{proof}
  This is clear, as the structure morphisms in $\Optic_\C$ are exactly the structure morphisms in $\C \times \C^\op$ under $\iota$.
\end{proof}

@@

In particular, for any $S \in \C$, the identity $\id_S$ yields an optic $c_S = \rep{\rho_S^{-1}}{\rho_S} : (S, S) \hto (I, I)$, that we will call the connector:
\begin{center}
  \input{diagrams/connector-full.tikz}
\end{center}
Or, again omitting the unitors:
\begin{center}
  \input{diagrams/connector.tikz}
\end{center}

@@

\begin{proposition}\label{prop-change-of-action-monoidal}
  A monoidal functor $F : \C \to \D$ induces a functor $\Optic(F) : \Optic_\C \to \Optic_\D$.
\end{proposition}
\begin{proof}
  On objects this is simply $\Optic(F)(S, S') = (FS, FS')$. On a morphism $\rep{l}{r} : (S, S') \to (A, A')$, we define
  \begin{align*}
    \Optic(F)\rep{l}{r} := \rep{\phi^{-1}_{M,A} (Fl)}{(Fr) \phi_{M,A'}}
  \end{align*}
  where $\phi_{M,A} : FM \otimes FA \to F(M \otimes A)$ is the structure map of the monoidal functor.
  \todo{functoriality}
  \todo{monoidality}
\end{proof}

\begin{proposition}\label{prop-optic-functor}
  Collectively this operation defines a 2-functor \[\Optic : \SymmMonCat \to \SymmMonCat.\] \qed
\end{proposition}
\todo{Haven't actually checked natural transformations}
\todo{Also we run into the large coend issue again, it's not defined on every symm mon cat.}

@@

In this section we establish a universal property of the $\Optic$ construction. The idea is that every optic $\rep{l}{r} : (S, S') \hto (A, A')$ consists of a morphism $S \to M \otimes A$ and the `formal dual' of a morphism $M \otimes A' \to S'$, composed with a `formal counit' that traces out the object $M$:
\begin{center}
  \input{diagrams/generic-optic-folded.tikz}
\end{center}

We assume in this section that $\C$ is a \emph{strict} symmetric monoidal category. It will be convenient to equip $\Optic_\C$ with a slightly different symmetric monoidal structure. The universal property for $\Optic_\C$ given in this section is an argument for this being the ``morally correct'' tensor, although it does seem a little strange.

\begin{definition}
  The \emph{switched} monoidal product on $\Optic_\C$ is given on objects by
  \begin{align*}
    (S, S') \switched (T, T') := (S \otimes T, T' \otimes S')
  \end{align*}
  And on morphisms $\rep{l}{r} : S \hto A$ and $\rep{l'}{r'} : T \hto B$ by:
  \begin{center}
    \input{diagrams/switched-tensor-on-morphisms.tikz}
  \end{center}
\end{definition}

@@

  The categories are monoidally equivalent via the identity functor \[I : (\Optic_\C, \switched, (I, I)) \to (\Optic_\C, \otimes, (I, I)),\] where the structure isomorphisms are given by
  \begin{align*}
    \iota(\id_{S \otimes S'}, s_{T, T'}) : I(S, S') \otimes I(T, T') \to I((S, S') \switched (T, T'))
  \end{align*}

@@

Note that because $\C_d$ is symmetric monoidal and has the same collection of objects as $\C$, the symmetric monoidal structure morphisms of $\C$ must be contained in $\C_d$.

@@

\begin{example}
  \todo{The funny graph example from the coherence paper?}
\end{example}

@@

  Note that the diagrams that are required to commute in the definition of teleological category all terminate with the unit $I$, so in view of Proposition~\ref{prop-costates} we should not be surprised that they correspond to equality of maps in $\C$.
\end{proof}

@@

  First, $F$ preserves the dualisable morphisms. If $\iota(f, g) : (S, S') \hto (A, A')$,
  \begin{align*}
    F\iota(f, g)
    &= F(\rep{\lambda_A^{-1} f}{g \lambda_{A'}}) \\
    &= \rep{\phi^{-1}_{I,A} (F(\lambda_A^{-1} f))}{(F(g \lambda_{A'})) \phi_{I,A'}} \\
    &= \rep{\phi^{-1}_{I,A} (F\lambda_A^{-1}) (Ff)}{(Fg)(F \lambda_{A'}) \phi_{I,A'}} \\
    &= \rep{(\phi^{-1}_I \otimes FA) \phi^{-1}_{I,A} (F\lambda_A^{-1}) (Ff)}{(Fg)(F \lambda_{A'}) \phi_{I,A'} (\phi_I \otimes FA)} \\
    &= \rep{\lambda_{FA}^{-1} (Ff)}{(Fg)\lambda_{FA'}} \\
    &= \iota(Ff, Fg)
  \end{align*}

  And it also preserves the counits:
  \begin{align*}
    F(\varepsilon_{(S, S')})
    &= F(c_{S \otimes S'}) \\
    &= F(\rep{\rho_S^{-1}}{\rho_{S'}}) \\
    &= \rep{\phi^{-1}_{S,I} (F \rho_S^{-1})}{(F \rho_{S'}) \phi_{S',I}} \\
    &= \rep{(FS \otimes \phi_I^{-1}) \phi^{-1}_{S,I} (F \rho_S^{-1})}{(F \rho_{S'}) \phi_{S',I} (FS \otimes \phi_I)} \\
    &= \rep{\rho_{FS}^{-1}}{\rho_{FS'} } \\
    &= \varepsilon_{(FS, FS')}
  \end{align*}

@@

To establish the universal property of $\Optic_\M$, we will

\begin{proposition}\label{prop-optic-decompose}
  Suppose $\rep{l}{r} : (S, S') \hto (A, A')$ has complement $M$. Then
  \begin{align*}
    \rep{l}{r} = ((A, I) \switched \varepsilon_{(M, I)} \switched (I, A'))(j(s_{M,A}l) \switched j(rs_{A',M})^*)
  \end{align*}
  where $j : \C \to \Optic_\C$ is the functor $j(A) := \iota(A, I)$.
\end{proposition}
The symmetries in the above expression could have been avoided if $\Optic$ had been defined as $\int^{M \in \C} \C(S, A \otimes M) \times \C(A' \otimes M, S')$, but it is too late to change the convention now!
\begin{proof}
  First note that because $\C$ is strict monoidal, the counit $\varepsilon_{(M, I)} : (M \otimes I, M \otimes I) = (M, M) \hto (I, I)$ is equal to the connector $c_M : (M, M) \hto (I, I)$.

  Then, up to strictness of the monoidal unit, we are composing the two optics
  \begin{center}
    \input{diagrams/optic-decomposed-outer.tikz}
  \end{center}
  and
  \begin{center}
    \input{diagrams/optic-decomposed-inner.tikz}
  \end{center}
  so the two pairs of twists cancel, and we are left exactly with the diagram for $\rep{l}{r}$.
\end{proof}

\begin{theorem}
  \label{optic-is-free-teleological-cat}
  Suppose $\C$ is a strict symmetric monoidal category and $\T$ is a teleological category. For all symmetric monoidal functors $F : \C \to \T_d$, there exists a teleological functor $K : \Optic_\C \to \T$, unique up to isomorphism, such that $Kj \cong F$.
\end{theorem}
\begin{proof}
  We begin by constructing $K$. On objects, define $K(S, S') = FS \otimes (FS')^*$. Suppose $\rep{l}{r} : (S, S') \hto (A, A')$ is an optic. By the previous Proposition,
  \begin{align*}
    \rep{l}{r} = ((A, I) \switched \varepsilon_{(M, I)} \switched (I, A'))(j(s_{M,A}l) \switched j(rs_{M,A'})^*)
  \end{align*}
  Which suggests defining
  \begin{align*}
    K\rep{l}{r} = (FA \otimes \varepsilon_{FM} \otimes (FA')^*)(F(s_{M,A}l) \otimes (F(rs_{A', M}))^* )
  \end{align*}
